### 2016-01-15 Documentation/systemic_risk.tex
newton_krylov

\documentclass[]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[utf8]{inputenc} 
\usepackage{subcaption}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{amsbsy}
\usepackage[left=4cm]{geometry}
 %http://tex.stackexchange.com/questions/595/how-can-i-get-bold-math-symbols
%opening
\title{}
\author{Pieter Van Nuffel}

\newcommand{\R}{\ensuremath{\mathbb{R}}} % commando zonder argumenten
\newcommand{\C}{\ensuremath{\mathbb{C}}}
\newcommand{\N}{\ensuremath{\mathbb{N}}}
\newcommand{\E}{\ensuremath{\mathbb{E}}}
\newcommand{\norm}[1]{\left\|#1\right\|} % commando's met argumenten
\newcommand{\pa}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\ppa}[2]{\frac{\partial^2 #1}{\partial #2^2}}
\newcommand{\dd}{\ensuremath{\mathrm{d}}}
\newcommand{\U}{\ensuremath{\boldsymbol{\rho}}}
\newcommand{\cts}{\ensuremath{\boldsymbol{\Phi}_T}} %Coarse time step
\newcommand{\V}{\ensuremath{\mathbf{v}}}
\newcommand{\jv}{\ensuremath{\mathbf{\hat{Jv}}}}
\newcommand{\jvpde}{\ensuremath{\mathbf{Jv}_{FP}}}

@@

We investigate the emergence of coarse equilibrium states in a stochastic system of p

 system of diffusion processes have a stabilizing force acting on each of them, corresponding to a bistable potential.

We develop a Newton-Krylov method that is able to compute accurately and efficiently coarse fixed points when the underlying fine-scale dynamics is stochastic. The main novelty of the algorithm is in the elimination of the noise that is generated when estimating Jacobian-vector products using time-integration of perturbed initial conditions. We present numerical results that demonstrate 
%the convergence properties of the numerical method, and use the method to show that macroscopic
%fronts in this model destabilise at a coarse symmetry-breaking bifurcation.
%
%We consider a system of diffusion processes that interact through their empirical mean and have a stabilizing force acting on each of them, corresponding to a bistable potential. There are three parameters that characterize the system: the strength of the intrinsic stabilization, the strength of the external random perturbations, and the degree of cooperation or interaction between them. The latter is the rate of mean reversion of each component to the empirical mean of the system. We interpret this model in the context of systemic risk and analyze in detail the effect of cooperation between the components, that is, the rate of mean reversion. We show that in a certain regime of parameters increasing cooperation tends to increase the stability of the individual agents but it also increases the overall or systemic risk. We use the theory of large deviations of diffusions interacting through their mean field.
%
%
%The present paper thus contains two main contributions. First, for the specific
%system under study, we explain the birth of the above-described macroscopic states in
%terms of coarse symmetry-breaking bifurcations. To the best of our knowledge, steps
%in this direction were taken only very recently [55, 7] and were confined to globally
%locked-in states. In the homogeneous case, we follow [5] and interpret metastable
%locked-in states as fixed points of a coarse evolution map. In the limit of infinitely
%many globally-coupled agents with homogeneous product preferences, we derive the
%coarse evolution map analytically. In the case of heterogeneous agents we employ
%stochastic continuation and show for the first time how fronts destabilise to partially
%locked-in states.
%The second main contribution of the paper is the development of a novel procedure
%to obtain coarse Jacobian-vector products with reduced variance, allowing
%the accurate evaluation of Jacobian-vector products in the presence of microscopic
%stochasticity, thus gaining full control over the linear and the nonlinear iterations
%of the Newton-Krylov solver. Even though our implementation of variance-reduced
%Jacobian-vector products is specific to the lock-in model, we believe that analogous
%strategies can be applied to other ABMs. Therefore, we provide a detailed account of
%the algorithmic steps involved in defining an accurate equation-free Newton-Krylov
%method and testing its convergence properties

@@

\section{Problem}

\subsection{Model problem}

\input{fokkerplanck_update.tex}

\subsection{Discretization}

\section{Method}

\subsection{Newton-Krylov solver} \label{sec:Newton-Krylov}

If we want to compute steady states for the density $\U_*$ without direct simulation, we can find them by solving the non-linear system

\begin{equation}
  {F}(\U_*) = \U_* - \cts(\U_*) =0. \label{non-linear_system}
\end{equation}
%where $\mathbf{F}$ is nonlinear function from $\R^N$ to  $\R^N$. 
To find the steady state $\U_*$, we apply Newton's method to eq. \ref{non-linear_system} . Starting from an initial state $\U^0$, we iterate

\begin{eqnarray}
\begin{cases}
& \texttt{Solve         }  J(\U^n) \boldsymbol{\delta_n}  =  - {F}(\U^n) \label{linear_system}        \\          
& \texttt{Set     } \U^{n+1} = \U^n+ \boldsymbol{\delta_n} 
\end{cases}
\end{eqnarray}
until convergence. 
$J(\U^n) =   F'(\U^n)  $ denotes the system Jacobian. Each Newton iteration $n$ thus involves evaluating the Jacobian of the timestepper $J(\cts(\U))$.
Since we do not have an explicit formula for $J(\cts)$, we are forced to use an iterative method, such as GMRES, that only requires Jacobian-vector products. %the action of the jacobian times a  vector
%$J(\cts)$, we are forced to use an iterative method, such as GMRES, that only requires Jacobian-vector products.
 \cite{Brown_Krylov}.
The Jacobian $J(\cts)$ applied to a vector $\mathbf{v}$ (with unit norm) will be estimated by a finite difference approximation
\begin{eqnarray}
\label{Jv_approx}
J(\cts) \cdot \mathbf{v} &\approx& \frac{\cts (\U + \varepsilon \mathbf{v}, \boldsymbol{\omega_1} )  - \cts (\U, \boldsymbol{\omega_2})}{\varepsilon} \\
&\approx & \frac{\cts (\U, \boldsymbol{\omega_1} )  + \varepsilon D(\cts) (  \U, \boldsymbol{\omega_1})  \cdot \mathbf{v}  - \cts (\U, \boldsymbol{\omega_2}) }{\varepsilon} \nonumber
.
\end{eqnarray}

\subsection{Variance reduced Jacobian-vector product}

\section{Analysis}
\subsection{Variance}
\subsection{bias}

\section{Application: Systemic Risk}

%\subsection{Including an interaction term}

Until now, we considered a system of non-interacting particles. We will now extend our model to a mean field model by introducing a third parameter $\alpha$, which is the degree of interaction or cooporation in the system. A simple form of cooperative behaviour is the case where each agent tends to follow the state of the majority (or, each particle feels an attractive force towards the mean state of the system). To include this cooperation effect in our stochastic simulation, we add this mean reversion term to the SDE \eqref{SDE}:

\begin{equation} 
\label{SDE_meanfield}
    \dd x = \mu V(x) \dd t + \sigma  \dd{W_t} + \alpha(\bar{x} -x) \dd t ,
\end{equation}
with  $\bar{x}(t) = \frac{1}{N} \sum_{i=1}^{N} x_i(t)$ denoting the empirical mean.

An interesting application of this model to banks and insurance is the emergence of systemic risk.  Banks will try to minimize their own individual risk by spreading the risk between each other. However, this may increase the risk that they may all fail: reducing individual risk on a micro-scale can increase systemic risk on a macro-scale. 
Garnier, Papanicolaou and Yang already made use of the dynamics in eq. \eqref{SDE_meanfield}  to show that interconnectedness between agents indeed affects the stability of the whole system, causing systemic risk  \cite{Garnier}. They defined $x_i$  as the state of risk of agent $i$. The bi-stable-state structure of the potential $V(x)$ ensures that each risk variable stays around $-1$ (defined as the normal state) or $+1$ (the failed state).  A natural measure of systemic risk is then the transition probability of the empirical mean $\bar{x}$ from the normal state to the failed state.

To establish the idea, let us repeat the numerical simulations with eq. \ref{SDE_meanfield}. The evolution of the system is now characterized by  by the initial conditions, the three parameters ($\mu$, $\sigma$, $\alpha$) and by the system size $N$. Figure \ref{fig:sysrisk}  illustrates the behavior of the empiral mean $\bar{x}$. The simulations were performed with all agents initially in the normal state. Nevertheless, if randomness dominates the interaction, the agents can move immediately to the other potential well. The system then behaves like $N$ independent difussions, and hence, by the symmetry of the potential, the mean state will be attracted to a single mixed state $\bar{x}=0$. Upon increasing the interaction parameter $\alpha$, however, we find two new macroscopic states, suggesting the precence of a pitchfork bifurcation at the macoscopic level. These solutions are no stable steady states, but rather coarse matastable states. Their lifetime is linked to the finite system size.

\begin{center}
\begin{figure}
\includegraphics[width=0.5\textwidth]{../../riskmodel/Problems/WeightedParticles/checkSystem/plots/xmean_alpha1}
\includegraphics[width=0.5\textwidth]{../../riskmodel/Problems/WeightedParticles/checkSystem/plots/xmean_alpha5}

\caption{The empirical mean, simulated for different $\alpha$. \textit{Left}: the system has one single state $\bar{x}=0$. \textit{Right}: for $\alpha> \alpha_c$ two metastable equilibria emerge. 
 The red dashed lines are approximated  analytical solutions for the steady states. \label{fig:sysrisk}}
\end{figure}
\end{center}

 \begin{center}
\begin{table}
\caption{Parameter values}
  \begin{tabular} { | c  c | c |}    \hline   
     %FTCS-scheme  &   Euler-Maruyama-scheme  \\  
   \textit{ {Discretization parameters}}&    &  SDE    \\ \hline
    Discretization step  & $\Delta x $  & $10^{-2}$ \\ 
        Number of discretization steps  & $n_x$ &  340 \\ 
    Time step  &  $\Delta t$ &   $10^{-2}$ \\ 
       Number of timesteps  & $n$  &  $10^{6}$ \\ \hline
  \end{tabular}
%\quad
%     \begin{tabular}  { | c | c | c |  }  \hline
%     \multicolumn{3}{|c|} {\textit{ System parameters}   }    \\
%    \hline  
%    Diffusion coefficient  & $D $ & $0.5$ \\ 
%   Drift coefficient &  $\mu$ & $ 1$ \\ \hline
%  \end{tabular}
\end{table}
\end{center}

\subsection{Bifurcation study}

\subsubsection{Calculating fixed points}
The steady  states are computed with the Newton-Krylov solver described in section \ref{sec:Newton-Krylov} and compared with the approximated analytical solutions, calculated by Garnier et. al for small $h$ \cite{Garnier}.

\begin{center}
\begin{figure}
\includegraphics[width=1\textwidth]{../../riskmodel/Problems/WeightedParticles/checkSystem/plots/steady_states.pdf}
\caption{\label{fig:bif_anal}}
\end{figure}
\end{center}

\subsubsection{Continuation}
We use a pseudo-arclength contunaution method with secant prediction steps.
In fig. \ref{todo} we show a bifurcation diagram of the densities.

\begin{center}
\begin{figure}
\includegraphics[width=0.8\textwidth]{../../riskmodel/Problems/WeightedParticles/checkSystem/plots/bif_alpha}
\caption{\label{fig:bif_anal}}
\end{figure}
\end{center}

%\begin{equation}
%\xi = sqrt(1-3*C)*(1+ 6*mu/(sigma**2)*C**2*(1-2*C)/(1-3*C))
%\end{equation}  
%
%

By adding the mean reversion term, the Fokker-Plack equation \eqref{fokkerplanck}  describing the evolution of the density, now becomes

\begin{equation}
\label{fp_mean_field}
\pa{\rho(x,t)}{t} = - \mu \pa{( V(x) \rho(x,t))}{x} - \alpha \pa{}{x} \left[ \left(\int x \rho(x,t) \mathrm{d}x  -x \right) \rho(x,t) \right] + \frac{\sigma^2}{2}  \ppa{\rho(x,t)}{x} .
\end{equation}

Explicit solutions of eq. \label{fp_mean_field} are not available in general, but we can find equilibrium solutions. Assuming

\bibliographystyle{plain}
\bibliography{biblio.bib}

>>>>

\documentclass[]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[utf8]{inputenc} 
\usepackage{subcaption}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{amsbsy}
\usepackage[left=4.5cm]{geometry}
%\usepackage{cancel}
 %http://tex.stackexchange.com/questions/595/how-can-i-get-bold-math-symbols
%opening
\title{}
\author{}

%\newcommand{\R}{\ensuremath{\mathbb{R}}} % commando zonder argumenten
\newcommand{\C}{\ensuremath{\mathbb{C}}}
\newcommand{\N}{\ensuremath{\mathbb{N}}}
\newcommand{\E}{\ensuremath{\mathbb{E}}}
\newcommand{\norm}[1]{\left\|#1\right\|} % commando's met argumenten
\newcommand{\pa}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\ppa}[2]{\frac{\partial^2 #1}{\partial #2^2}}
\newcommand{\dd}{\ensuremath{\mathrm{d}}}
\newcommand{\U}{\ensuremath{\boldsymbol{\rho}}}
\newcommand{\cts}{\ensuremath{\boldsymbol{\Phi}_T}} %Coarse time step
\newcommand{\V}{\ensuremath{\mathbf{v}}}
\newcommand{\jv}{\ensuremath{\mathbf{\hat{Jv}}}}
\newcommand{\jvpde}{\ensuremath{\mathbf{Jv}_{FP}}}
%\newcommand{\expect}[1]{\ensuremath { \mathbb{E} \left[ #1  \right] }

@@

%
%We investigate coarse equilibrium states of a stochastic model 
%
%
% we derive an analytic
%approximate coarse evolution-map for the expected average purchase. We then study the emergence
%of coarse fronts when the agents are split into two factions with opposite preferences. We
%develop a novel Newton-Krylov method that is able to compute accurately and efficiently coarse
%fixed points when the underlying fine-scale dynamics is stochastic. The main novelty of the algorithm
%is in the elimination of the noise that is generated when estimating Jacobian-vector products
%using time-integration of perturbed initial conditions. We present numerical results that demonstrate
%the convergence properties of the numerical method, and use the method to show that macroscopic
%fronts in this model destabilise at a coarse symmetry-breaking bifurcation.
%
%We consider a system of diffusion processes that interact through their empirical mean and have a stabilizing force acting on each of them, corresponding to a bistable potential. There are three parameters that characterize the system: the strength of the intrinsic stabilization, the strength of the external random perturbations, and the degree of cooperation or interaction between them. The latter is the rate of mean reversion of each component to the empirical mean of the system. We interpret this model in the context of systemic risk and analyze in detail the effect of cooperation between the components, that is, the rate of mean reversion. We show that in a certain regime of parameters increasing cooperation tends to increase the stability of the individual agents but it also increases the overall or systemic risk. We use the theory of large deviations of diffusions interacting through their mean field.
%
%
%The present paper thus contains two main contributions. First, for the specific
%system under study, we explain the birth of the above-described macroscopic states in
%terms of coarse symmetry-breaking bifurcations. To the best of our knowledge, steps
%in this direction were taken only very recently [55, 7] and were confined to globally
%locked-in states. In the homogeneous case, we follow [5] and interpret metastable
%locked-in states as fixed points of a coarse evolution map. In the limit of infinitely
%many globally-coupled agents with homogeneous product preferences, we derive the
%coarse evolution map analytically. In the case of heterogeneous agents we employ
%stochastic continuation and show for the first time how fronts destabilise to partially
%locked-in states.
%The second main contribution of the paper is the development of a novel procedure
%to obtain coarse Jacobian-vector products with reduced variance, allowing
%the accurate evaluation of Jacobian-vector products in the presence of microscopic
%stochasticity, thus gaining full control over the linear and the nonlinear iterations
%of the Newton-Krylov solver. Even though our implementation of variance-reduced
%Jacobian-vector products is specific to the lock-in model, we believe that analogous
%strategies can be applied to other ABMs. Therefore, we provide a detailed account of
%the algorithmic steps involved in defining an accurate equation-free Newton-Krylov
%method and testing its convergence properties

@@

\section{Systemic risk}

\begin{equation}
\label{fokkerplanck}
\pa{\rho(x,t)}{t} = - \mu \pa{( U(x) \rho(x,t))}{x} - \alpha \pa{}{x} \left[ \left(\int x \rho(x,t) \mathrm{d}x  -x \right) \rho(x,t) \right] + D  \ppa{\rho(x,t)}{x} 
\end{equation}

\begin{figure}
\includegraphics{../../riskmodel/Problems/WeightedParticles/checkSystem/plots/xmean_alpha5}

\end{figure}
